---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

# Installing Libraries

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
```


# Loading in Data for Each Month

```{r warning=FALSE}
jan_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic01-22/202201-divvy-tripdata.xls")

feb_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic02-22/202202-divvy-tripdata.xls")

mar_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic03-22/202203-divvy-tripdata.xls")

apr_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic04-22/202204-divvy-tripdata.xls")

may_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic05-22/202205-divvy-tripdata.xls")

jun_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic06-22/202206-divvy-tripdata.xls")

jul_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic07-22/202207-divvy-tripdata.xls")

aug_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic08-22/202208-divvy-tripdata.xls")

sep_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic09-22/202209-divvy-publictripdata.xls")

oct_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic10-22/202210-divvy-tripdata.xls")

nov_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic11-22/202211-divvy-tripdata.xls")

dec_data <- read_excel("/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/cyclistic12-22/202212-divvy-tripdata.xls")
```

# Check Structure of Each Month's Data to Ensure Consistencies
```{r}
str(jan_data)

str(feb_data)

str(mar_data)

str(apr_data)

str(may_data)

str(jun_data)

str(jul_data)

str(aug_data)

str(sep_data)

str(oct_data)

str(nov_data)

str(dec_data)
```


# Convert all non-numerical columns to character 
```{r}
jan_data <-  mutate(jan_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

feb_data <-  mutate(feb_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

mar_data <-  mutate(mar_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))  

apr_data <-  mutate(apr_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

may_data <-  mutate(may_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

jun_data <-  mutate(jun_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

jul_data <-  mutate(jul_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

aug_data <-  mutate(aug_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

sep_data <-  mutate(sep_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

oct_data <-  mutate(oct_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

nov_data <-  mutate(nov_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 

dec_data <-  mutate(dec_data, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 
```

# Merge Datasets

```{r}
year2022_data <- bind_rows(jan_data, 
                       feb_data, 
                       mar_data, 
                       apr_data, 
                       may_data, 
                       jun_data, 
                       jul_data, 
                       aug_data, 
                       sep_data, 
                       oct_data, 
                       nov_data, 
                       dec_data
                       )
```

# Inspecting Full Year Data

```{r}
str(year2022_data)
summary(year2022_data)
```

# Verifying that there are two unique values for member_casual 

```{r}
unique(year2022_data$member_casual)
```





# Creating columns for day, month, and year for aggregation purposes

```{r}
year2022_data$date <- as.Date(year2022_data$started_at)
year2022_data$month <- format(as.Date(year2022_data$date), "%m")
year2022_data$day <- format(as.Date(year2022_data$date), "%d")
year2022_data$year <- format(as.Date(year2022_data$date), "%Y")
```

# Creating new ride_length column that shows seconds

```{r}
year2022_data$ride_length <- difftime(year2022_data$ended_at,year2022_data$started_at)
```

# Inspecting structure of updated data

```{r}
str(year2022_data)
```

# Converting ride_length to numerical for easier calculations

```{r}
year2022_data$ride_length <- as.numeric(as.character(year2022_data$ride_length))
```

# Removing observations with incomplete data

```{r}
originalrows <- nrow(year2022_data)
year2022_data <- na.omit(year2022_data)
originalrows - nrow(year2022_data)
```
## 47480 rows with missing data were removed

# Descriptive Analysis

## Calculate Summary Statistics of Ride Length
```{r}
summary(year2022_data$ride_length)
```

## Subtract Observations with Negative Trip Length

```{r}
year2022_data <- year2022_data[!(year2022_data$ride_length < 0),]
```

## New Summary Statistics

```{r}
summary(year2022_data$ride_length)
```

## Comparing Members vs. Casual Riders

```{r}
aggregate(year2022_data$ride_length ~ year2022_data$member_casual, FUN = mean)
aggregate(year2022_data$ride_length ~ year2022_data$member_casual, FUN = median)
aggregate(year2022_data$ride_length ~ year2022_data$member_casual, FUN = max)
aggregate(year2022_data$ride_length ~ year2022_data$member_casual, FUN = min)
```

## Average Ride Time by Day of Week for Members vs. Casual Riders
```{r}
aggregate(year2022_data$ride_length ~ year2022_data$member_casual + year2022_data$day_of_week, FUN = mean)
```

## Average Ride Time by Month for Members vs. Casual Riders

```{r}
aggregate(year2022_data$ride_length ~ year2022_data$member_casual + year2022_data$month, FUN = mean)
```

## Visualization of Number of Rides by Rider Type

```{r}
year2022_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

## Visualization for Average Duration

```{r}
year2022_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

# Exporting CSV File to Further Analyze Average Ride Length

```{r}
counts <- aggregate(year2022_data$ride_length ~ year2022_data$member_casual + year2022_data$day_of_week, FUN = mean)

write.csv(counts, file = 'C:/Users/kheng/OneDrive/Documents/Google Career Certificate Case Study/cyclisticData/avg_ride_length.csv')
```